---
title: "Final project"
editor: visual
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(ggplot2)
library(scales)
library(dplyr)
library(tidyr)
```

# Internal Revenue Service (IRS) Exempt Organizations Division: Factors and variables that influence revenue and income in specific regions.

## Abstract

## Data Set

| Variables        | Descriptions                              |
|------------------|-------------------------------------------|
| EIN              | Employer Identification number            |
| Name             | Primary name                              |
| ICO              | In care of Name                           |
| Street           | Street address                            |
| City             | City                                      |
| State            | State                                     |
| Zip              | Zip Code                                  |
| Group            | Group Exemption Number                    |
| Subsection       | Subsection code                           |
| Affiliation      | Affiliation Code                          |
| Classification   | Class code                                |
| Ruling           | Ruling Date                               |
| Deductibility    | Deductibility code                        |
| Foundation       | Foundation Code                           |
| Activity         | Activity Code                             |
| Organization     | Organization code                         |
| Status           | EO status code                            |
| Tax_period       | Tax period                                |
| Asset_CD         | Asset code                                |
| Income_cd        | Income code                               |
| Filing_req_CD    | Filing requirement code                   |
| Pf_Filing_req_CD | PF filing requirement code                |
| ACCT_PD          | Accounting period                         |
| Asset_AMT        | Asset amount                              |
| Income_AMT       | Income amount                             |
| Revenue_AMT      | revenue amount                            |
| NTEE_CD          | National Taxonomy of exempt Entities code |
| Sort_Name        | Sort name                                 |

```{r}

mid.atlantic <- read.csv("tax.2024.csv", stringsAsFactors = FALSE, na.strings = c("", "NA"))
num_cols <- c("ASSET_AMT", "INCOME_AMT", "REVENUE_AMT")

mid.atlantic[num_cols] <- lapply(mid.atlantic[num_cols], function(x) as.numeric(x))
mid.atlantic_clean <- mid.atlantic %>%
  filter(!is.na(INCOME_AMT),
         INCOME_AMT > 0,
         !is.na(NTEE_CD),
         NTEE_CD != "")
mid.atlantic_clean <- mid.atlantic_clean %>% filter(REVENUE_AMT > 0)
tax2024<-mid.atlantic_clean

str(mid.atlantic_clean)



```

data set description

## How does an organization’s NTEE classification (mission: health, environment) relate to income?

```{r}
mid.atlantic_clean %>%
  group_by(NTEE_CD) %>%
  summarise(median_income = median(INCOME_AMT, na.rm = TRUE)) %>%
  slice_max(order_by = median_income, n = 20) %>%   # top 20 groups
  ggplot(aes(x = reorder(NTEE_CD, median_income), y = median_income)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 20 NTEE Codes by Median INCOME_AMT",
       x = "NTEE Code",
       y = "Median Income") +
  theme_minimal()



```

```{r}
mid.atlantic_clean %>%
  group_by(NTEE_CD) %>%
  summarise(median_revenue = median(REVENUE_AMT, na.rm = TRUE))  %>%
  slice_max(order_by = median_revenue, n = 20)%>%
  ggplot(aes(x = reorder(NTEE_CD, median_revenue), y = median_revenue)) +
  geom_col() +
  coord_flip() +
  labs(title = "Median REVENUE_AMT by NTEE Code",
       x = "NTEE Code",
       y = "Median Revenue") +
  theme_minimal()

```

```         
```

## What organizational characteristics predict high revenue nationally?

mlr

```{r}

model_mlr <- lm(
  log(REVENUE_AMT + 1) ~ 
    ASSET_AMT + INCOME_AMT + ASSET_CD + INCOME_CD,
  data = mid.atlantic_clean
)

summary(model_mlr)

```

### Which organizational characteristics most strongly predict whether a nonprofit is high-revenue?

```{r}
tax2024$HighIncome <- factor(tax2024$HighIncome, levels = c("Low", "High"))

# Create NTEE major group (first letter)
tax2024$NTEE_major <- substr(tax2024$NTEE_CD, 1, 1)
tax2024$log_assets <- log1p(tax2024$ASSET_AMT)

tax2024$GroupStatus <- factor(tax2024$GroupStatus)
tax2024$STATE <- factor(tax2024$STATE)
model_org_category <- glm(
  HighIncome ~ NTEE_major + GroupStatus + STATE +
    ASSET_AMT + ACCT_PD,
  data = tax2024,
  family = binomial
)
summary(model_org_category)
```

```{r}
set.seed(123)
index <- sample(1:nrow(tax2024), 0.8*nrow(tax2024))
train <- tax2024[index, ]
test  <- tax2024[-index, ]

model_test <- glm(
  HighIncome ~ NTEE_major + GroupStatus + STATE +
    ASSET_AMT + ACCT_PD,
  data = train,
  family = binomial
)

pred <- ifelse(predict(model_test, test, type = "response") > 0.5, "High", "Low")
table(pred, test$HighIncome)
mean(pred == test$HighIncome)
```
