---
title: "Final project"
editor: visual
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(ggplot2)
library(scales)
library(dplyr)
library(tidyr)
library(randomForest)
tax2024 <- read.csv("tax.2024.csv", stringsAsFactors = FALSE, na.strings = c("", "NA"))
num_cols <- c("ASSET_AMT", "INCOME_AMT", "REVENUE_AMT")

tax2024[num_cols] <- lapply(tax2024[num_cols], function(x) as.numeric(x))
tax2024 <- tax2024 %>%
  filter(!is.na(INCOME_AMT),
         INCOME_AMT > 0,
         !is.na(NTEE_CD),
         NTEE_CD != "")
tax2024  <- tax2024 %>% filter(REVENUE_AMT > 0)
tax2024  <- tax2024  %>% filter(INCOME_AMT > 0)


str(tax2024)
```

## Introduction

## Question 1

#### Which organizational characteristics most strongly predict whether a nonprofit is high-revenue?

To determine how factors such as mission type (NTEE major group), organizational structure (group-affiliated vs standalone), geography (state), financial attributes (assets, revenue), and fiscal year-end month contribute to predicting whether a nonprofit has income above the median for the year 2024.

Logistic regression

```{r}
tax2024$HighIncome <- factor(tax2024$HighIncome, levels = c("Low", "High"))

# Create NTEE major group (first letter)
tax2024$NTEE_major <- substr(tax2024$NTEE_CD, 1, 1)


tax2024$GroupStatus <- factor(tax2024$GroupStatus)
tax2024$STATE <- factor(tax2024$STATE)
model_org_category <- glm(
  HighIncome ~ NTEE_major + GroupStatus + STATE + ACCT_PD,
  data = tax2024,
  family = binomial
)
summary(model_org_category)
anova(model_org_category, test = "Chisq")
set.seed(123)
index <- sample(1:nrow(tax2024), 0.8*nrow(tax2024))
train <- tax2024[index, ]
test  <- tax2024[-index, ]

model_test <- glm(
  HighIncome ~ NTEE_major + GroupStatus + STATE + ACCT_PD,
  data = train,
  family = binomial
)

pred <- ifelse(predict(model_test, test, type = "response") > 0.5, "High", "Low")
table(pred, test$HighIncome)
mean(pred == test$HighIncome)
```

random forest

```{r}
set.seed(456) 
model_rf <- randomForest(
  HighIncome ~ NTEE_major + GroupStatus + STATE + ACCT_PD,
  data = train,
  ntree = 200,        
  importance = TRUE   
)


pred_rf <- predict(model_rf, test)


table(pred_rf, test$HighIncome)
mean(pred_rf == test$HighIncome)
```

```         
•   Logistic Regression
•   Decision Trees
•   Random Forest
•   KNN
```

<br>

## Question 2

#### To what extent does geography (state-level location) influence whether a nonprofit is high-revenue?

Test whether revenue patterns vary significantly by state and whether geography is a meaningful predictor after controlling for other factors.

```         
•   Logistic Regression
•   Decision Tree
•   Random Forest
```

<br>

## Question 3

#### Does mission type (NTEE major group) significantly influence whether a nonprofit is high-revenue?

Focus on how mission category (Arts, Education, Health, Human Services, Religion, etc.) contributes to financial performance.

```         
•   Logistic Regression
•   Tree / Random Forest
•   KNN (optional)
```

<br>

## Question 4

#### Do group-affiliated nonprofits differ from standalone nonprofits in their likelihood of being high-revenue?

Examine whether belonging to a parent organization (group ruling) affects financial capacity

```         
•   Logistic Regression
•   Decision Tree
•   Random Forest<br>
```

## Question 5

#### Can a nonprofit’s mission type (NTEE major group) be predicted using its financial indicators and organizational attributes?

Flip the previous questions: instead of predicting income from mission, predict mission from financial structure.

```         
•   Multinomial Logistic Regression
•   Random Forest
•   KNN
```

<br>

## Question 6

#### Are organizational characteristics able to predict whether a nonprofit has a high asset-to-revenue ratio?

Look at financial efficiency or stability instead of raw income.

Same classification models.

<br>

## Key Findings

<br>

## Future Scope

<br>
